Equation-free analysis of a dynamically evolving multigraph

### 2016-02-05 report/multigraph_manuscript.tex
Moved to Newton, added figs to manuscript
Added new default param values to ./coarse_ng based on what has worked
in the cpi routine. still quite unstable. no longer allow cpi to
continue if a projection failed.

\documentclass[12pt]{article}
\usepackage{graphicx, subcaption, setspace, indentfirst, amsfonts, amsthm, empheq}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[center]{titlesec}
\graphicspath{ {./figs/paper/} }
\pagestyle{plain}
\renewcommand\thesection{\Roman{section}}
\renewcommand\thesubsection{\alph{subsection}.}
\newcommand\ignore[1]{}
\setlength{\parskip}{0ex}
\def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
\let\endchangemargin=\endlist

\title{\vspace{-5mm}Equation-free analysis of a dynamically evolving multigraph}
\author{Alexander Holiday,$^1$ Bal\'{a}zs R\'{a}th,$^2$ Ioannis G. Kevrekidis$^{1,3}$ \\ \\
  $^1$\textit{Department of Chemical and Biological Engineering}, \\
  \textit{Princeton University, Princeton, New Jersey 08544, USA} \\ \\
  $^2$\textit{Department of Mathematics}, \\
  \textit{The University of British Columbia, Vancouver, British Columbia V6T 1Z4}
}
\date{}

@@

\begin{abstract}
In order to adapt traditional techniques from numerical analysis to complex network systems, we apply the equation-free framework to a dynamically evolving multigraph system. This approach is based on the coupling of short intervals of simulation with appropriately-defined lifting and restriction operators that map the full network description to suitable macroscopic variables, enabling both the acceleration of direct simulations through coarse-projective integration, and the identification of stationary states via a Newton-GMRES method. Additionally, we employ the dimensionality-reduction algorithms principal component analysis (PCA) and diffusion maps (DMAPS) to uncover hidden structure in the model.
\end{abstract}

@@

Over the past decades, myriad systems have been formulated as complex evolving networks, from highway traffic \cite{joubert_large-scale_2010} to brain connectivity \cite{hermundstad_learning_2011}. When modeling such systems, dynamics are typically defined at a very fine scale, specifying individual node and edge interactions; explicit closed equations governing macroscopic properties are often unavailable \cite{durrett_graph_2012, joubert_large-scale_2010, roche_agent-based_2011, swaminathan_modeling_1998}. These interactions are often complicated functions dependent on multiple system parameters, which, when coupled with the large network sizes inherent in many interesting problems, makes numerical investigation computationally prohibitive. However, the lack of explicit equations governing macroscopic property evolution precludes the use of traditional numerical techniques such as optimization and fixed-point finding that could offer valuable insight into the network's behavior, leaving little alternative but to work with the full network system. Faced with this dilemma, investigators must either restrict their attention to a smaller, heuristically-motivated parameter space \cite{hodgkin_quantitative_1952} or simplify the model, potentially removing important features \cite{brown_variability_1999}. \par

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{step-illustration}
  \caption{Illustration of the steps enumerated above that evolve the graph from $t$ to $t+1$ \label{fig:step-illustration}}
\end{figure}

@@

\begin{figure}[ht!]
  % TODO
  \centering
  \includegraphics[width=0.5\textwidth]{triangle-slaving}
  \caption{Figure showing evolution of triangle counts over time,
    starting from three different initial networks that shared the
    same degree sequence. Thus higher-order variables that may be
    initialized off the slow manifold are quickly drawn to it
    ($O(n^2)$ steps). \label{fig:sv}}
\end{figure}

@@

The lifting operator performs the inverse role of restriction: translating coarse variables into full networks. Specifically in the context of our model, we map vectors of polynomial coefficients to full networks in a two-stage process. First, the coefficient vector is used to recreate a sorted degree sequence by evaluating the full polynomial at integer values and rounding to the nearest degree, as depicted in Fig. (\ref{fig:cpi-diagram}). If the sum of the resulting degree sequence is odd, a single degree is added to the largest value to ensure the sequence is graphical. Second, this degree sequence is used as input to a Havel-Hakimi algorithm, creating a graph whose degree sequence matches that of the input \cite{havel_remark_1955, hakimi_realizability_1962}. While the canonical Havel-Hakimi method produces simple graphs, it is not difficult to extend it to multigraphs by allowing the algorithm to wrap around the sequence, producing multiple edges and self-loops. The lifting procedure is represented by the red arrows of Fig. (\ref{fig:cpi-diagram}).

@@

\begin{figure}[ht!]
  % TODO
  \centering
  \includegraphics[width=0.5\textwidth]{placeholder}
  \caption{Evolution of a single polynomial coefficient over time, both with and without CPI to show agreement. \label{fig:coeff-evo}}
\end{figure}

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{cpi-error}
    \subcaption{Percent error for each vertex in a CPI-accelerated simulation compared to an ensemble of plain simulations. \label{fig:cpi-error}}
  \end{subfigure} \\ %
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{self-error}
    \subcaption{Percent error when comparing two ensembles of plain runs, showing the discrepancies arising simply from the stochastic nature of the system. \label{fig:self-error}}
  \end{subfigure}%
  \caption{Figure showing small errors in CPI routine. \label{fig:cpi-results}}
\end{figure}

@@

When applying machine learning techniques to a dataset, it is necessary to define a distance (or similarity) between each data point. If these points are a set of vectors in $\mathbb{R}^n$ one has a number of metrics to choose from, the Euclidean distance being a common first option. Unfortunately, when individual points are not vectors but networks, the definition of a computationally feasible metric becomes far more challenging. Examples such as the maximal common subgraph and edit distances, defined in \cite{bunke_graph_1998} and \cite{gao_survey_2010} do define metrics on the space of graphs, but their computation is $NP-hard$. \\

@@

\begin{figure}[ht!]
  % TODO
  \centering
  \includegraphics[width=0.75\textwidth]{pca-2d-rho-kappa}
  \caption{Embedding of steady states in $\rho$ / $\kappa$ plane,
    showing that PCA can uncover the two-dimensional structure of the
    dataset. \label{fig:pca}}
\end{figure}

@@

Unlike PCA, DMAPS uncovers underlying nonlinear manifolds hidden in data. This is achieved by solving for the discrete equivalent of the eigenfunctions and eigenvalues of the Laplacian operator over the manifold, which amounts to calculating leading eigenvector/eigenvalue pairs of a symmetric Markov matrix $W$. See \cite{coifman_diffusion_2006, nadler_diffusion_2006} for details. Results from embeddings on different timescales are shown in Fig. (\ref{fig:dmaps-results}). We see that on the short, $n^2$ timescale, the differences in initial conditions lead to well-separated embeddings in three-dimensions represented by the small spheres in Fig. (\ref{fig:dmaps-results}), while at long times, we see that both systems evolve to the same steady state over a shared one-dimensional curve, as would be expected given the one-dimensional parameterization possible in time. \\

@@

\begin{figure}[ht!]
  % TODO
  \centering
  \includegraphics[width=0.75\textwidth]{dmaps-2d-rho-kappa}
  \caption{Embedding of steady states in $\rho$ / $\kappa$ plane,
    showing that DMAPS uncovered the coarse variables known from
    Bal\'{a}zs' paper \label{fig:dmaps-rk}}
\end{figure}

>>>>

\documentclass[epjST, final]{svjour}
\usepackage{graphicx, subcaption, setspace, indentfirst, amsfonts, amsthm, empheq}
% \usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
% \usepackage[center]{titlesec}
\graphicspath{ {./figs/paper/} }
% \pagestyle{plain}
% \renewcommand\thesection{\Roman{section}}
% \renewcommand\thesubsection{\alph{subsection}.}
\newcommand\ignore[1]{}
% \setlength{\parskip}{0ex}
% \def\changemargin#1#2{\list{}{\rightmargin#2\leftmargin#1}\item[]}
% \let\endchangemargin=\endlist

\title{Equation-free analysis of a dynamically evolving multigraph}
\author{Alexander Holiday \inst{1} \and Bal\'{a}zs R\'{a}th \inst{2}
  \and Ioannis G. Kevrekidis \inst{1.3}} \institute{Department of
  Chemical and Biological Engineering, Princeton University,
  Princeton, New Jersey 08544, USA \and Institute of Mathematics,
  Budapest University of Technology, 1 Egry J\'{o}zsef u., Budapest
  1111, Hungary \and PACM, Princeton University, Princeton, New Jersey
  08544, USA}
% $^1$\textit{Department of Chemical and Biological Engineering},
% \textit{Princeton University, Princeton, New Jersey 08544, USA} \\ \\
% $^2$\textit{Department of Mathematics}, \\
% \textit{The University of British Columbia, Vancouver, British
%   Columbia V6T 1Z4} } \date{}

\abstract{In order to adapt traditional techniques from numerical
  analysis to complex network systems, we apply the equation-free
  framework to a dynamically evolving multigraph system. This
  approach is based on the coupling of short intervals of simulation
  with appropriately-defined lifting and restriction operators that
  map the full network description to suitable macroscopic
  variables, enabling both the acceleration of direct simulations
  through coarse-projective integration, and the identification of
  stationary states via a Newton-GMRES method. Additionally, we
  employ the dimensionality-reduction algorithms principal component
  analysis (PCA) and diffusion maps (DMAPS) to uncover hidden
  structure in the model.}

@@

@@

Over the past decades, myriad systems have been formulated as complex evolving networks, from highway traffic \cite{joubert_large-scale_2010} to brain connectivity \cite{hermundstad_learning_2011}. When modeling such systems, dynamics are typically defined at a very fine scale, specifying individual node and edge interactions; explicit closed equations governing macroscopic properties are often unavailable \cite{durrett_graph_2012,joubert_large-scale_2010,roche_agent-based_2011,swaminathan_modeling_1998}. These interactions are often complicated functions dependent on multiple system parameters, which, when coupled with the large network sizes inherent in many interesting problems, makes numerical investigation computationally prohibitive. However, the lack of explicit equations governing macroscopic property evolution precludes the use of traditional numerical techniques such as optimization and fixed-point finding that could offer valuable insight into the network's behavior, leaving little alternative but to work with the full network system. Faced with this dilemma, investigators must either restrict their attention to a smaller, heuristically-motivated parameter space \cite{hodgkin_quantitative_1952} or simplify the model, potentially removing important features \cite{brown_variability_1999}. \par

@@

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{graph-evo}
  \caption{Illustration of the steps enumerated above that evolve the graph from $t$ to $t+1$ \label{fig:step-illustration}}
\end{figure}

@@

\begin{figure}[ht!]
  % TODO
  \centering
  \includegraphics[width=0.8\textwidth]{triangle-slaving}
  \caption{Figure showing evolution of triangle counts over time,
    starting from three different initial networks that shared the
    same degree sequence. Thus higher-order variables that may be
    initialized off the slow manifold are quickly drawn to it
    ($O(n^2)$ steps). \label{fig:sv}}
\end{figure}

@@

The lifting operator performs the inverse role of restriction: translating coarse variables into full networks. Specifically in the context of our model, we map vectors of polynomial coefficients to full networks in a two-stage process. First, the coefficient vector is used to recreate a sorted degree sequence by evaluating the full polynomial at integer values and rounding to the nearest degree, as depicted in Fig. (\ref{fig:cpi-diagram}). If the sum of the resulting degree sequence is odd, a single degree is added to the largest value to ensure the sequence is graphical. Second, this degree sequence is used as input to a Havel-Hakimi algorithm, creating a graph whose degree sequence matches that of the input \cite{havel_remark_1955,hakimi_realizability_1962}. While the canonical Havel-Hakimi method produces simple graphs, it is not difficult to extend it to multigraphs by allowing the algorithm to wrap around the sequence, producing multiple edges and self-loops. The lifting procedure is represented by the red arrows of Fig. (\ref{fig:cpi-diagram}).

@@

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{cpi-3d-comp}
    \subcaption{Percent error for each vertex in a CPI-accelerated simulation compared to an ensemble of plain simulations. \label{fig:cpi-error}}
  \end{subfigure} \\ %
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{cpi-error-comp}
    \subcaption{Percent error when comparing two ensembles of plain runs, showing the discrepancies arising simply from the stochastic nature of the system. \label{fig:self-error}}
  \end{subfigure}%
  \caption{Figure showing small errors in CPI routine. \label{fig:cpi-results}}
\end{figure}

@@

When applying machine learning techniques to a dataset, it is necessary to define a distance (or similarity) between each data point. If these points are a set of vectors in $\mathbb{R}^n$ one has a number of metrics to choose from, the Euclidean distance being a common first option. Unfortunately, when individual points are not vectors but networks, the definition of a computationally feasible metric becomes far more challenging. Examples such as the maximal common subgraph and edit distances, defined in \cite{bunke_graph_1998} and \cite{gao_survey_2010} do define metrics on the space of graphs, but their computation is $NP\textnormal{-}hard$. \\

@@

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{pca-2d-rho}
    \subcaption{Coloring the two-dimensional PCA embedding with
      $\rho$. \label{fig:pca-rho}}
  \end{subfigure} \\ %%
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{pca-2d-kappa}
    \subcaption{Coloring the two-dimensional PCA embedding with
      $\kappa$. \label{fig:pca-kappa}}
  \end{subfigure}%
  \caption{Embedding of steady states in $\rho$ / $\kappa$ plane,
    showing that PCA can uncover the two-dimensional structure of the
    dataset. \label{fig:pca}}
\end{figure}

@@

Unlike PCA, DMAPS uncovers underlying nonlinear manifolds hidden in data. This is achieved by solving for the discrete equivalent of the eigenfunctions and eigenvalues of the Laplacian operator over the manifold, which amounts to calculating leading eigenvector/eigenvalue pairs of a symmetric Markov matrix $W$. See \cite{coifman_diffusion_2006,nadler_diffusion_2006} for details. Results from embeddings on different timescales are shown in Fig. (\ref{fig:dmaps-results}). We see that on the short, $n^2$ timescale, the differences in initial conditions lead to well-separated embeddings in three-dimensions represented by the small spheres in Fig. (\ref{fig:dmaps-results}), while at long times, we see that both systems evolve to the same steady state over a shared one-dimensional curve, as would be expected given the one-dimensional parameterization possible in time. \\

@@

\begin{figure}[h!]
  \vspace{-5mm}
  \centering
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{dmaps-2d-rho}
    \subcaption{Coloring the two-dimensional DMAPS embedding with
      $\rho$. \label{fig:dmaps-rho}}
  \end{subfigure} \\ %%
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=3in]{dmaps-2d-kappa}
    \subcaption{Coloring the two-dimensional DMAPS embedding with
      $\kappa$. \label{fig:dmaps-kappa}}
  \end{subfigure}%
  \caption{Embedding of steady states in $\rho$ / $\kappa$ plane,
    showing that DMAPS uncovered the variables governing the
    stationary state, known from
    Bal\'{a}zs' paper \label{fig:dmaps-rk}}
\end{figure}
